Mega-analysis of the Interoceptive Accuracy Scale (IAS) Structure and its Subjective Correlates

### 2025-10-13 paper/1_Initial/manuscript.qmd
updated manuscript

The original validation study suggested a two-factor structure for the IAS, one reflecting the perception of general interoceptive signals (urinate, hungry, defecate, thirsty, pain, heart, taste, breathing, temperature, muscles, affective touch, vomit, sexual arousal), and other relating to signals that may be difficult to perceive solely through interoceptive information (itch, tickle, cough, burp, bruise, blood sugar, sneeze, wind). 
The authors however underlined its acceptable but imperfect fit [@murphy2019, p. 127], and several follow-up studies have indeed identified different optimal solutions. 
For instance, @brand2023 reported a 1-factor solution, while @lin2023 - using Exploratory Graph Analysis [EGA, @golino2017exploratory] - and @campos2021 found bifactor solutions [i.e., one general factor above a set of lower-level factors, @rodriguez2016evaluating] to be the best fit. 
Using a 2-factors Exploratory Factor Analysis (EFA), @koike2023 suggested that the items could be grouped into cutaneous (itching, tickling, coughing, burping, affective touch, bruising, passing gas, sneezing, muscle sensations, sexual arousal, and taste) and visceral sensations (urination, defecation, hunger, thirst, pain, breathing, fatigue/blood sugar, temperature, vomiting, and heartbeat).

@@

<!-- EGA -->
By combining network analysis with psychometric methods, the recently-developed EGA framework allows to jointly estimate the number of dimensions (i.e., groups of items), the structure as well as its stability [@golino2017exploratory; @golino2020investigating]. 
Evidence has underlined its suitability as an alternative to traditional factor analysis, addressing some of its limitations such as the assumption of a "latent" source of variability, possible biasing in the estimation of the optimal factor numbers depending on sample size, and the poor performance of other methods in complex population structures, while remaining comparable and interpretable [@christensen2021equivalency; @jimenez2023dimensionality]. 
At a fundamental level, EGA conceptualizes variables as nodes in a network, with connections (edges) reflecting associations between them. 
Clustering these nodes reveals distinct communities of related items, in practice akin to traditional latent factors - but without explicitly assuming their presence [@christensen2021equivalency]. 
We used the EGAnet package [@christensen2021estimating] to fit a hierachical EGA with the Leiden community detection algorithm.

@@

<!-- AIms -->
The present study aimed to provide a comprehensive evaluation of the IAS, spanning both its structure (study 1) and its dispositional correlates (study 2). Together, these studies offer novel insights into the psychometric characteristics, conceptual interpretation, and practical applications of the IAS within the broader landscape of interoception research.
<!-- Study 1 findings -->
Study 1 demonstrated that the IAS is best conceptualised as clusters of tightly coupled item pairs, reflecting discrete bodily domains such as visceral sensations, musculoskeletal cues, and expulsion phenomena. The pairs emerged once several items (e.g., Tickle, Taste, Affective Touch) were removed due to their instability, redundancy or poor associations with the other items. The resulting 14-item refined IAS retained the psychometric integrity of the original scale while improving interpretability and efficiency.

A short version of the IAS has been proposed before. @lin2023 suggested a 12-item version by removing items with local dependency, high difficulty, and/or low discrimination. Their parallel analysis suggested unidimensionality, although this method is known to underestimate factor numbers in multidimensional data [@markos2024dimensionality;@golino2017exploratory]. In contrast, our study applied multiple methods, yet the same pairs consistently emerged across datasets and demonstrated superior fit compared to alternative models.This challenges the assumption of unidimensionality [@lin2023; @murphy2019] and resonates with theoretical perspectives that describe interoception as a multimodal, context-sensitive process rather than a singular capacity (@khalsa2018; @vlemincx2023novel].

<!-- Study 2 findings -->
Study 2 provided further evidence for a domain-specific perspective. Certain item pairings, particularly Hunger–Thirst and Bruise–Blood sugar, showed consistent negative correlations with depression and anxiety measures, in some cases approaching or even surpassing the strength of associations observed for the full IAS scale. Notably, these were also the only pairs, out of the seven examined, that displayed significant positive correlations with the BPQ Body Awareness, whereas the full IAS scale showed no such association. This pattern suggests that specific item pairings may show distinct patterns of association with other interoceptive measures and with mental health indices, whereas the aggregated IAS scale may obscures these relationships. This highlights that questionnaires like the IAS may capture multiple facets of interoception rather than a single underlying dimension, and that domain-specific approaches may provide greater construct validity.

These findings mirror broader inconsistencies in the literature concerning associations between the IAS and the BPQ-A. While some studies report significant negative associations [@lin2023], others have observed significant positive associations [@gaggero2021; @todd2022; @brand2022; @campos2021], and still others none at all [@murphy2019]. In our study the full IAS was not associated with the BPQ-A. Interpretation is further complicated by conceptual debates: the BPQ Body Awareness is frequently treated as a measure of interoception, whereas the BPQ Autonomic Reactivity subscale has been criticized as indexing dysfunction in the nervous system rather than interoceptive sensitivity [@gaggero2021]. Our results lend tentative support to this interpretation, as the full IAS scale and most item pairings showed significant negative associations with the BPQ Autonomic Reactivity, with the exception of the Bruise–Blood sugar pair. This pattern is consistent with the idea that this subscale may capture maladaptive bodily experiences or autonomic dysfunction, rather than interoceptive accuracy per se. However, the Body Awareness subscale of the BPQ has also been criticized for its ambiguous instructions. It is often unclear whether respondents should interpret the items as assessing attentional focus on bodily sensations, perceived accuracy in detecting them, or simply the frequency of experiencing such signals. As a result, studies have reported that observed associations vary depending on how the BPQ-A is construed [@campos2021; @gaggero2021; @gabriele2022dissociations]. As such, it is unclear whether the lack of a relationship between the IAS and the BPQ-A reflects a genuine divergence between the constructs measured by the two scales, or instead arises from ambiguities in how the BPQ-A is operationalized and interpreted across studies.

Taken together with these debates, the discrepancies observed here highlight that some IAS pairings may capture interoceptive self-reported accuracy more effectively than the aggregated measure, raising questions about the interpretability of the full scale. By contrast, other pairings (e.g., Cough–Sneeze, Wind–Burp) were weakly or inconsistently related to external variables, raising concerns about their usefulness. Together, these findings indicate that the strength and relevance of interoceptive self-reported accuracy may vary systematically across physiological domains. At the same time, the interpretation of such associations is complicated by the lack of contextualisation in the items themselves. For example, anxiety is often experienced as heart racing (ref), yet similar sensations occur during exercise or heightened arousal more generally. Without contextual anchors, it remains unclear which bodily states participants are drawing upon when responding. In line with prior work emphasizing the importance of contextual factors in the measurement of interoception [@vlemincx2023novel], our results suggest that the value of self-report lies less in broad aggregation across domains and more in the targeted assessment of physiologically and situationally meaningful contexts.

<!-- Correlates of IAS Across Psychological Domains -->
The dispositional correlates of the IAS provide important conceptual boundaries for the construct. Interoceptive accuracy, as assessed by the IAS, was robustly associated with mood-related outcomes, particularly depression and anxiety, underscoring its relevance for affective regulation and bodily state monitoring [@brand2023; @lin2023]. This is in line research consistently revealing the link between conditions, such as depression or anxiety and interoceptive measures. For example, divergent findings in subjective and objective measures of interoception have been linked with high anxiety symptomology [@garfinkel2016interoceptive]. At the same time, recent studies have reported null or inconsistent associations between interoceptive abilities and mental health symptomatology, suggesting that such links may not generalize across all interoceptive domains or measurement approaches [ref pre-print?]. This further highlights the need to examine interoceptive self-perception at the level of separate domains rather than aggregated scores.

In contrast, associations of IAS scores with personality traits and higher-order beliefs were modest: small positive correlations emerged with Openness and Agreeableness, and weak associations with primal world beliefs. Consistent with recent work linking interoceptive self-report to personality dispositions [@haustein2024interoception;@simkute2025interoception], these results suggest that interoceptive accuracy primarily operates at the level of bodily self-perception. This contrast reinforces the view that the IAS primarily indexes bodily self-perception, exerting only modest influence on broader dispositional traits. A potential key reason for these discrepancies is the lack of a consistent theoretical framework for interoception, as dimensions are neither defined nor captured equally across questionnaires and experimental tasks [@desmedt2022measures].[**complete paragraph**]

@@

<!-- Limitations -->
One limitation worth noting is that we did not evaluate the IAS alongside other widely used questionnaires, such as the BPQ or MAIA. Although both are frequently employed in interoception research, they present important limitations. For example, the MAIA-2 contains subscales (e.g., Not-Worrying, Not-Distracting) with weak reliability and minimal relation to a general interoception factor [@ferentzi2021examining; @rogowska2023validation]. Several other subscales primarily reflect higher-level cognitive tendencies, such as mindfulness and attentional regulation, rather than interoceptive body awareness [@vig2023self; @todd2022]. Similarly, the BPQ’s Body Awareness and Autonomic Reactivity subscales show conceptual and psychometric concerns, and BPQ scores often overlap substantially with negative affect, raising questions about their specificity as measures of interoceptive sensibility [@campos2021; @gabriele2022dissociations; @vig2023self].

@@

<!-- Future Outlook in the field -->

Building on these limitations, a next-generation interoception scale should balance breadth (multiple items per modality, including underrepresented domains such as cardiac and respiratory) with depth (domain-specific subscales). Retaining analogue or wider rating ranges (e.g., 1–7 instead of 1–5) would further enhance sensitivity and coverage. Such a scale would improve the interpretability, sensitivity, and practical utility of interoceptive self-report measures, ultimately allowing for a more nuanced mapping of interoceptive abilities to affective, cognitive, and health-related outcomes.

>>>>

The original validation study suggested a two-factor structure for the IAS, one reflecting the perception of general interoceptive signals (urinate, hungry, defecate, thirsty, pain, heart, taste, breathing, temperature, muscles, affective touch, vomit, sexual arousal), and other relating to signals that may be difficult to perceive solely through interoceptive information (itch, tickle, cough, burp, bruise, blood sugar, sneeze, wind). 
The authors however underlined its acceptable but imperfect fit [@murphy2019, p. 127], and several follow-up studies have indeed identified different optimal solutions. 
For instance, @brand2023 reported a 1-factor solution, while @lin2023 and @campos2021 found bifactor solutions [i.e., one general factor above a set of lower-level factors, @rodriguez2016evaluating] to be the best fit. 
Using a 2-factors Exploratory Factor Analysis (EFA), @koike2023 suggested that the items could be grouped into cutaneous (itching, tickling, coughing, burping, affective touch, bruising, passing gas, sneezing, muscle sensations, sexual arousal, and taste) and visceral sensations (urination, defecation, hunger, thirst, pain, breathing, fatigue/blood sugar, temperature, vomiting, and heartbeat).

@@

<!-- EGA -->
By combining network analysis with psychometric methods, the recently-developed Exploratory Graph Analysis [EGA, @golino2017exploratory] framework allows to jointly estimate the number of dimensions (i.e., groups of items), the structure as well as its stability [@golino2017exploratory; @golino2020investigating]. 
Evidence has underlined its suitability as an alternative to traditional factor analysis, addressing some of its limitations such as the assumption of a "latent" source of variability, possible biasing in the estimation of the optimal factor numbers depending on sample size, and the poor performance of other methods in complex population structures, while remaining comparable and interpretable [@christensen2021equivalency; @jimenez2023dimensionality]. 
At a fundamental level, EGA conceptualizes variables as nodes in a network, with connections (edges) reflecting associations between them. 
Clustering these nodes reveals distinct communities of related items, in practice akin to traditional latent factors - but without explicitly assuming their presence [@christensen2021equivalency]. 
We used the EGAnet package [@christensen2021estimating] to fit a hierachical EGA with the Leiden community detection algorithm.

@@

<!-- AIms -->
The present study aimed to provide a comprehensive evaluation of the IAS, spanning both its structure (study 1) and its dispositional correlates (study 2). Together, these studies offer novel insights into the psychometric characteristics, conceptual interpretation, and practical applications of the IAS within the broader landscape of interoception research.

<!-- Study 1 findings -->
Study 1 demonstrated that the IAS is best conceptualised as clusters of tightly coupled item pairs, reflecting discrete bodily domains such as visceral sensations, musculoskeletal cues, and expulsion phenomena. The pairs emerged once several items (e.g., Tickle, Taste, Affective Touch) were removed due to their instability, redundancy or poor associations with the other items. The resulting 14-item refined IAS retained the psychometric integrity of the original scale while improving interpretability and efficiency.

A short version of the IAS has been proposed before. @lin2023 suggested a 12-item version by removing items with local dependency, high difficulty, and/or low discrimination. Their parallel analysis suggested unidimensionality, although this method is known to underestimate factor numbers in multidimensional data [@golino2017exploratory; @markos2024dimensionality]. In contrast, our study applied multiple methods such as EGA, FA, and HCA yet the same pairs consistently emerged across datasets and demonstrated superior fit compared to alternative models. This challenges the assumption of unidimensionality of the IAS [@murphy2019; @brand2023]. Although the short version of the IAS by @lin2023 preferred a one-factor solution, their 12-item IAS was derived by removing locally dependent items identified through Item Response Theory (IRT). In our framework, these locally dependent items are instead interpreted as meaningful pairs (e.g., Heart–Breath), reflecting distinct bodily domains. However, because IRT assumes unidimensionality [@nguyen2014introduction], it may have treated these natural item clusters as sources of statistical dependency rather than as theoretically relevant structures within interoceptive accuracy.

<!-- Study 2 findings -->
Study 2 provided further evidence for a domain-specific perspective. Certain item pairings, particularly Hunger–Thirst and Bruise–Blood sugar, showed consistent negative correlations with depression and anxiety measures, in some cases approaching or even surpassing the strength of associations observed for the full IAS scale. Notably, these were also the only pairs, out of the seven examined, that displayed significant positive correlations with the BPQ Body Awareness, whereas the full IAS scale showed no such association. This pattern suggests that specific item pairings may show distinct patterns of association with other interoceptive measures and with mental health indices, whereas the aggregated IAS scale may obscure these relationships. This highlights that questionnaires like the IAS may capture multiple facets of interoception rather than a single underlying dimension, and that domain-specific approaches may provide greater construct validity.

These findings mirror broader inconsistencies in the literature concerning associations between the IAS and the BPQ Body Awareness. While some studies report significant negative associations [@lin2023], others have observed significant positive associations [@brand2023; @campos2021; @gaggero2021; @todd2022], and still others none at all [@murphy2019]. In our study the full IAS was not associated with the BPQ Body Awareness. Interpretation is further complicated by conceptual debates: the BPQ Body Awareness is frequently treated as a measure of interoception, whereas the BPQ Autonomic Reactivity subscale has been criticized as indexing dysfunction in the nervous system rather than interoceptive sensitivity [@gaggero2021]. Our results lend tentative support to this interpretation, as the full IAS scale and most item pairings showed significant negative associations with the BPQ Autonomic Reactivity, with the exception of the Bruise–Blood sugar pair. This pattern is consistent with the idea that this subscale may capture maladaptive bodily experiences or autonomic dysfunction, rather than interoceptive accuracy per se. However, the Body Awareness subscale of the BPQ has also been criticized for its ambiguous instructions. It is often unclear whether respondents should interpret the items as assessing attentional focus on bodily sensations, perceived accuracy in detecting them, or simply the frequency of experiencing such signals. As a result, studies have reported that observed associations vary depending on how the BPQ Body Awareness is constructed [@campos2021; @gabriele2022dissociations; @gaggero2021]. As such, it is unclear whether the lack of a relationship between the IAS and the BPQ Body Awareness in this study reflects a genuine divergence between the constructs measured by the two scales, or instead arises from ambiguities in how the BPQ Body Awareness is operationalized and interpreted across studies.

Taken together with these debates, the discrepancies observed here highlight that some IAS pairings may capture interoceptive self-reported accuracy more effectively than the aggregated measure, raising questions about the interpretability of the full scale. By contrast, other pairings (e.g., Cough–Sneeze, Wind–Burp) were weakly or inconsistently related to external variables, raising concerns about their usefulness. Together, these findings indicate that the strength and relevance of interoceptive self-reported accuracy may vary systematically across physiological domains. At the same time, the interpretation of such associations is complicated by the lack of contextualisation in the items themselves.For example, a racing heart is often associated with anxiety, yet the same sensation can also arise during exercise or other forms of physiological arousal. Without contextual anchors, it remains unclear which bodily states participants are drawing upon when responding. In line with prior work emphasizing the importance of contextual factors in the measurement of interoception [@vlemincx2023novel], our results suggest that the value of self-report lies less in broad aggregation across domains and more in the targeted assessment of physiologically and situationally meaningful contexts.

<!-- Correlates of IAS Across Psychological Domains -->
The dispositional correlates of the IAS help delineate the conceptual boundaries of interoceptive accuracy as a construct. Self-reported interoceptive accuracy, as assessed by the IAS, has been robustly associated with mood-related outcomes, particularly depression and anxiety, underscoring its relevance for affective regulation and bodily state monitoring (Benau et al., 2023; Brand et al., 2023). This aligns with a broader body of research linking affective conditions such as anxiety and depression with altered interoceptive processing (e.g., Garfinkel et al., 2016). However, recent findings have also reported null or inconsistent associations between interoceptive abilities and mental health symptomatology, suggesting that such relationships may not generalize across all interoceptive domains or measurement modalities [@banellis2025interoceptive; @jenkinson2024interoception]. For instance, a recent review by @jenkinson2024interoception found no consistent evidence linking cardiac interoceptive accuracy tasks with anxiety or depression across multiple meta-analyses. Collectively, these findings underscore the importance of examining interoceptive self-perception at the level of distinct bodily domains rather than relying on aggregated or unidimensional measures.

In contrast, associations of IAS scores with personality traits and higher-order beliefs were modest: small positive correlations emerged with Openness and Agreeableness, and weak associations with primal world beliefs. Consistent with recent work linking interoceptive self-report to personality dispositions [@haustein2024interoception; @simkute2025interoception], these results suggest that interoceptive accuracy primarily operates at the level of bodily self-perception. This contrast reinforces the view that the IAS primarily indexes bodily self-perception, displaying only modest associations with  broader dispositional traits.

A potential key reason for these discrepancies is the lack of a consistent theoretical framework for interoception, as dimensions are neither defined nor captured equally across questionnaires and experimental tasks [@desmedt2022measures]. **[complete paragraph]**

@@

<!-- Limitations -->
One limitation worth noting is that we did not evaluate the IAS alongside other widely used questionnaires, such as the BPQ or MAIA. Although both are frequently employed in interoception research, they present important limitations. For example, the MAIA-2 contains subscales (e.g., Not-Worrying, Not-Distracting) with weak reliability and minimal relation to a general interoception factor [@ferentzi2021examining; @rogowska2023validation]. Several other subscales primarily reflect higher-level cognitive tendencies, such as mindfulness and attentional regulation, rather than interoceptive body awareness [@todd2022; @vig2023self]. Similarly, the BPQ’s Body Awareness and Autonomic Reactivity subscales show conceptual and psychometric concerns, and BPQ scores often overlap substantially with negative affect, raising questions about their specificity as measures of interoceptive sensibility [@campos2021; @gabriele2022dissociations; @vig2023self].

@@

<!-- Future Outlook in the field -->
Building on these limitations, a next-generation interoception scale should balance breadth (multiple items per modality, including underrepresented domains such as cardiac and respiratory) with depth (domain-specific subscales). Retaining analogue or wider rating ranges (e.g., 1–7 instead of 1–5) would further enhance sensitivity and coverage. Such a scale would improve the interpretability, sensitivity, and practical utility of interoceptive self-report measures, ultimately allowing for a more nuanced mapping of interoceptive abilities to affective, cognitive, and health-related outcomes.
